Material Methods

### 2020-03-04 doc/MatMeth.Rmd

Todo

Add the single Bivalent predictions to one of the sections.

@@

To access natural genetic variation for *Mus musculus*, wild derived inbred strains were used. The Mus musculus strains PWD/PhJ, PERC/Eij, WSB/EiJ, LEWE/EiJ, MSM/MsJ, MOLF/EiJ, CAST/EiJ, CZECHII/EiJ were purchased from Jackson labs (Maine, USA). The strains of KAZ/TUA, TOM/TUA, AST/TUA, and HMI/TUA were generated from frozen embryos purchased  from Biological Resource Center (BRC) at Riken (Ibaraki, Japan).  (https://en.brc.riken.jp) and rederived by the UW-Biotech centre (Madison UW).
The related murid species *Mus caroli* CAROLI/EiJ and *Mus spretus* SPRET/EiJ were purchased from Jackson labs (https://www.jax.org) and *Mus spicilegus* SPIC/Eij from Riken.

 All mice were housed UW-Madison Biotech and MSC facilities - following the protocols.  A breeding colony of wild derived *Mus musculus* mice sampled from Gough Island (GI)is maintained at UW Veterinary school facilities.
 Mice were fed on dry standard breeder chow. Some strains sunflower seeds, nestlets and larger cages were added to improve fertility and litter survival. Adult mice were euthanized by CO asphyxiation. Neonate and embryonic mice were euthanized by decapitation following the protocols approved by the Institutional Animal Care and Use Committee at the University of Wisconsin-Madison.

Over the course of data collection some breeding colonies  mice were moved from facilities. Additionally the GI strain was kept at a separate facility.  (evidence for environmental (room) effect on (MLH1 / SC lengths / a variety of meiotic traits was tested for. Effects on average MLH1 counts were compared -- for strains which were kept at mulitple institutions. ((We observed no effect, Supplemental figure))

@@

Spermatocyte spreads were collected and perpared as described in [@peterson_2019] . The majority of mice used were between 5 and 12 weeks, (supplemental table). However some strains had problems breeding, so testes from older mice were collected. We found a (small/no) age effect on MLH1 counts for male mice (Supplemental Figure)).

The only difference between preparation of spermatocyte and oocyte spreads was the volumns of the hypotonic buffer was 300 u l instead of 3ml.

The vast majority of oocyte data was collected from neonate litters/mice between 5 to 48 hours old. This approach maintain breeding pairing and still result in prophase oocytes (cite timeline of oocytes in neonates). (differences were assessed between embryo samples and neonates?).

Precise staging of embryos via copulary plugs was not feasible in many of these wild derived strains due to their behavior. Embryonic samples were collected when pregnancy noted in females, embryos were staged based on (X table markers).

Meiocyte spreads were made following [@peters_1997]. For testes the tunica was removed and whole testes was incubated in 3ml of hypotonic solution for 45min. For ovaries after dissection, the pair of ovaries were decapsulated in cold PBS and both were incubated in 300ul of hypotonic solution for 45 min.

After incubation, gonads were transferred to sucrose solution -for mastication to make a cell slurry which was transferred to 80ul of 2% PFA solution. Cells were fixed in this solution and dried in a humid chamber at room temperature overnight. The next morning, slides were treated with a photoflow (Kodak, cite) wash.

Staining / Immunohistochemistry

SYCP3 was used to mark  visualize the Axial element of the synaptonemal complex. Sicne all mouse chromosomes are telocentric/acrocentric, CREST was used to indicate the centromere -- indicate the side of the bivalent with the centromere.

the genome wide recombination rate was charaecterized / estimated with staining of MLH1 antibody, 
The same staining protocol was applied to spermatocyte and oocyte spreads.
The staining protocol was based on X with some adaptations, and same as previously described in [@peterson_2019].

Precursurs to COs, DSBs were quantified with the DMC1 staning antibody. Due to resource constraints -- only a subset of strains were stained for this marker. Only spermatocyte spreads produced enough good cells for DMC1 quantification.

@@

Images of cells were capture using a Zeiss Axioplan 2 microscope with AxioLab camera and Axio Vision software (Zeiss, Cambridge, UK).

Preprocessing, including cropping, noise reduction, and histogram adjustments, was performed using Photoshop (v13.0).

For MLH1 pachytene characterization, cells with a full karyotype (19 acrocentric bivalents and XY for spermatocytes or 20 acrocentric bivalents for oocytes), distinct foci, and intact bivalents were included for quantification.
Image file names were anonymized before manual scoring of MLH1 or DMC1 foci.

@@

Hand measures of meiotic chromosomes generated by using ImageJ/Fiji (v1.52) [@fiji].

@@

Foci counts/numbers were manually scored.

@@

DMC1 -- Ttest

MLH1 counts per cell are good /decent proxies for genome wide recombination rates estimates (cite).

@@

The mean, variance and coefficient of variance for MLH1 counts per cell were calculated for each mouse (in R). these mouse level summary statistics were analyzed -- within our model frameworks.

@@

We can also compare the size of the sex effects between (subspecies) and within (strain) mouse groups. (ie is the male or female strain effect larger? is there more variance due to random strain effect in male or females?)
(we used this frame work as proxy for 'polymorphism' and 'divergence' in MLH1 counts across house mouse and closely related species.)
I may have to do more research/reading on testing the random effects.

### Single bivalet quant

Single bivalent chromosomes were isolated from cytology cell images as discribed in (X ). the single bivalents were 'curated' (removed errors in the segmentation algorithm)

to account for effects of CO number and physical chromosome size -- the data was also analyzed after spliting up by chromosome class and in long and short chromosome bins.

- Split up 'classes' of bivalents depending on the number of foci -- to tests/ comparisons more comprable.

@@

- whole cell bivalent measures (measures of all bivalents ), were compiled by mixed automated bivalent measures with hand measurements. (Since the error in automated measures was so low)

**SC length**

The axial element (AE) of the SC -- for single bivalents (synapsesd homolog pairs), was quantified as the skeletonized (single pixel width line) --

-comparison to human measures previously in permyscus species -- and in musculuc (shown in the supplement)

@@

**CO interference**

- how we choose to measure interference, the average distance between2 foci on same bivalent.

- downside to using the gamma distribution -- not the primary meteric for CO interference.

- Also pooling intervals for categories and fitting a gamma distribution (in supplement)

**rec landscape** **1CO normalized position**

- how why measure the recombination landscape -- coohse to charaecterize the rec landscaoe bby isloating the 1CO bivalents; since there are known sexual dimorphic features of this and since bivalents with more than 1 foci will be primarily influenced by CO interference. We look at both the raw and

the distance to the telomere and centromere end of the SC shape were measured -- for all / pooled bivalents

-specific R packages
-non-parametric  (tests and measures for dealing with count data)
-hypothesis tested (size, DSB number)

-Mixed models and glm models were built

-tests for fixed effects
-tests of random effects in mixed models was LTR of exactRLRT function in X package.

single bivalent analysis

# Numbers of mice cells ect for each method?

above belongs more in results

**outline**

1. framework for analyzing the MLH1 evolutionary patterns across mus species complex. (what is the main question we were tackling when designing this framework?)

2. write equation, describe terms (proxies for within and between polymorphism)
logic / choices for fixed and random effects

3. what are the 2 general / basic questions this framework can be applied to.

$X \sim mouse average MLH1 ~ sex + subspecies + random(strain) + \xi $

4. how can DMC1 measures be added to the mixed model

5. descriptions of mlh1 outside of framework (missing female data)

>>>>

@@

To access natural genetic variation for *Mus musculus*, wild derived inbred strains were used. The _Mus musculus_ strains PWD/PhJ, PERC/Eij, WSB/EiJ, LEWE/EiJ, MSM/MsJ, MOLF/EiJ, CAST/EiJ, CZECHII/EiJ were purchased from Jackson labs (Maine, USA). The strains of KAZ/TUA, TOM/TUA, AST/TUA, and HMI/TUA were generated from frozen embryos purchased  from Biological Resource Center (BRC) at Riken (Ibaraki, Japan).  (https://en.brc.riken.jp) and rederived by the UW-Biotech centre (Madison UW).
The related murid species *Mus caroli* CAROLI/EiJ and *Mus spretus* SPRET/EiJ were purchased from Jackson labs (https://www.jax.org) and *Mus spicilegus* SPIC/Eij from Riken. We use abriviated superscript notation for the remainder of the paper (e.g. _musculus^WSB^_).

 All mice were housed UW-Madison Biotech and MSC facilities following RARC approved protocols.  A breeding colony of wild derived _Mus musculus_ mice sampled from Gough Island (GI)is maintained at UW Veterinary school facilities.
 Mice were fed on dry standard breeder chow. Some strains sunflower seeds, nestlets and larger cages were added to improve fertility and litter survival. Adult mice were euthanized by CO asphyxiation. Neonate and embryonic mice were euthanized by decapitation following the protocols approved by the Institutional Animal Care and Use Committee at the University of Wisconsin-Madison. Over the course of data collection some breeding colonies  mice were moved from facilities. Additionally the GI strain was kept at a separate facility. We tested for effects on mean MLH1 counts across difference mouse rooms/facilities and found no effect (Supplemental section).

@@

Spermatocyte spreads were collected and perpared as described in [@peterson_2019] . The majority of mice used were between 5 and 12 weeks, (supplemental table). for DMC1 spreads we targeted juvinile male mice (X ages).

Because some strains have breeding issues, all mice some strains were kept in breeding pairs (before collecting, and aged out of the desired age range. We performed tests for effects on MLH1 counts and found no effects in the older male mice ages. (Also we tested other effects, including maternal age...)

The vast majority of oocyte data was collected from neonate mice between 5 to 48 hours old. This approach maintain breeding pairing and still result in prophase oocytes <(cite timeline of oocytes in neonates)>. . Precise staging of embryos via copulary plugs was not feasible in many of these wild derived strains due to their behavior. Embryonic samples were collected when pregnancy noted in females, embryos were staged based on (X table markers). No effect on mlh1 count due to age effects (pooled as embryo and neonate) was found (Supplemental section).

Meiocyte spreads were made following [@peters_1997].  whole testes and decapsulated overaires were incubated in 3ml or 300ul of hypotonic solution for 45min respectively.

After incubation, gonads were transferred to sucrose solution for mastication. 15ul of cell slurry was transferred to 80ul of 2% PFA solution. Cells were fixed in this solution and dried in a humid chamber at room temperature overnight. The next morning, slides were treated with a photoflow (Kodak, cite) wash.

**Staining / Immunohistochemistry**

SYCP3 was used to mark  visualize the Axial element of the synaptonemal complex. Sicne all mouse chromosomes are telocentric/acrocentric, CREST was used to indicate the centromere <insert the concentrations>

Precursurs to COs, DSBs were quantified with the DMC1 staning antibody.

@@

```{r error.repeatbility, echo=FALSE}

```

Images of cells were capture using a Zeiss Axioplan 2 microscope with AxioLab camera and Axio Vision software (Zeiss, Cambridge, UK). Preprocessing, including cropping, noise reduction, and histogram adjustments, was performed using Photoshop (v13.0).

For MLH1 pachytene quantification, cells with a full karyotype (19 acrocentric bivalents and XY for spermatocytes or 20 acrocentric bivalents for oocytes), distinct foci, and intact bivalents were included for quantification.
Image file names were anonymized before manual scoring of MLH1 or DMC1 foci.

@@

@@

(the data from these scripts differed slightly in that the total sc script returned the skeletonized SC area for the whole cell. <state the script parameters, (morphogy (3) and other parameter)>

to quantify / charaecterize the SC (length) (synapsed chromosomes, )-- two image analysis algorithms were used both which apply a 'skeletonizing' transformation to the synapsed chromosomes,  which transforms the elongatde chromosome space into a single pixel wide 'trace' of the originial chromosom shape.

The DNA CrossOver script isolated single chromosomes shapes from an image, in addition to returning the SC length -- also measured the location of (green and blue signal along the single SC - reflecting MLH1 foci and cetromere signal respectively).

Hand measures of SC/pactyene / meiotic chromosomes generated by using ImageJ/Fiji (v1.52) [@fiji].

@@

@@

The mean, variance and coefficient of variance for MLH1 counts per cell were calculated for each mouse (in R).

**M1. Mixed Model** $$mouse \ av.\ MLH1\ foci\ per\ cell ~=~ subsp * sex + rand(strain) + \varepsilon $$

**M2. Linear Model** $$mouse \ av.\ MLH1\ foci\ per\ cell ~=~ subsp * sex * strain + \varepsilon$$

**M3. Linear Model** $$mouse \ av.\ MLH1\ foci\ per\ cell ~=~ sex * strain + \varepsilon$$

**M4. Linear Model** $$sex\ specific\ mouse \ av.\ MLH1\ foci\ per\ cell ~=~ subsp * strain + \varepsilon$$

@@

### Single bivalent charaecterization

@@

The single bivalent data set was curated of algorithm mistakes manually>. For charaecterizing any of the foci positions or interfocal distances any normalized refers to dividing the raw measures by the total SC for that invidiualy bivalent.

The proportions of different chromosome classes (seperated by number of foci) was tested with chi-sqaure tests. <in R using the X package/function>

**chromosome size effects**

Using the single bivalent dataset, the within cell distribution of bivalent SC lengths from cells with at least 17 isolated chromosome measures. The longest and shortest 4 to 5 chromosomes within each cell were identified by qualqulating the fourth and second quartille. (these comprised the short and long bivalent datasets).

@@

< manual measures of whole cell bivalent measures (measures of all bivalents were incorporated into this data set. ), were compiled by mixed automated bivalent measures with hand measurements. (Since the error in automated measures was so low)>

**CO Interferecne**

Our primary mode of examining interference is to compared the interfocal disatnces (IFDs) (raw and normalized by SC length.)
Gamma parameter v has historically been used to charaecterize interference and was calculated using <x function>.
